---
title: "Coding challanges for Session Five"
subtitle: ""
author: "Akos Mate"
date: '2021 March'
output:
    html_document:
        code_folding: "show"
        number_sections: FALSE
        toc: true
        toc_depth: 4
        toc_float: true
        theme: flatly
        highlight: tango
        df_print: paged
        css: ../rmd_style.css
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      message = FALSE,
                      warning = FALSE
)
```


```{r echo=FALSE}

library(tidyverse)

```


We are using some new datasets for practice. If you are interested in various datasets for practice or personal pet projects, a good starting point is the [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday) GitHub repo where they post a new dataset each week.


# Excercise 1 - Practicing `ggplot`

> Using the 'squads' data, create a scatterplot where the number of caps (international games played) is depicted along the X-axis, goals (international goals scored) is visualized at the Y-axis. Let's add colors based on the footballer's position.

More info on the dataset: https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-07-09

```{r}
squads <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-07-09/squads.csv")
```



```{r echo=FALSE}
library(ggplot2)

ggplot(data = squads,
       aes(x = caps,
           y = goals,
           color = pos)) +
  geom_point()

```




> Now, overlay multiple histograms on each other -- use fill = variable to have a histogram of goals with players of different position depicted with different colors. You might want to trim (set the limits of) the x-axis at 75 goals, as very few players score more than that. 

```{r echo=FALSE}
ggplot(data = squads,
       mapping = aes(x = goals,
                     fill = pos)) +
  xlim(0, 75) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.65)

```

> Create a histogram of caps (international games played), fill the bars with grey color, and overlap it with a density plot, filled with the color 'salmon'! 

```{r echo=FALSE}
ggplot(squads,
       mapping = aes(x = caps)) +
  geom_histogram(aes(y = ..density..),
                 binwidth = 1,
                 fill = "grey",
                 color = "black") +
  xlim(0, 200) +
  geom_density(alpha = 0.25, fill = "salmon")
```



Now we use the other part of the World Cup data.

```{r}
wwc_outcomes <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-07-09/wwc_outcomes.csv")
```

> Create a joyplot of *scores* using the wwc_outcomes data, using the result (i.e. whether the team won, lost or tied) as a grouping variable! 

```{r}
library(ggridges)
```

```{r echo=FALSE}
ggplot(data = wwc_outcomes,
       mapping = aes(x = score,
                     y = win_status,
                     fill = win_status)) +
  geom_density_ridges(scale = 0.8, alpha = 0.5)
```

