---
title: "Packages to install"
subtitle: "Session Three"
author: "Akos Mate"
date: '2020 July'
output:
    html_document:
        code_folding: "show"
        number_sections: TRUE
        toc: true
        toc_depth: 4
        toc_float: true
        theme: flatly
        highlight: tango
        css: ../rmd_style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      message = FALSE
)
```


Since we covered how to install packages, please run the following script in your RStudio Cloud project to install all the neccesary packages that we'll use in the course. This will speed things up considerably. If you run into any problems during the install, drop us a line on Slack.

```{r eval=FALSE}

packages_needed <- c("tidyr", "ggridges", "ggthemes", "ggraph", "tidygraph", "eurostat", "maps", "ggfortify", "effects", "GGally", "broom", "purrr")

install.packages(packages_needed)

```


For a dataset we'll use, we need to install the package from GitHub, with the following code

```{r eval=FALSE}
install.packages("remotes")
remotes::install_github("allisonhorst/palmerpenguins")
```

