---
title: "Writing functions and flow control in R"
author: "Akos Mate"
subtitle: "Session Five"
date: '2022 February'
output:
    html_document:
        code_folding: "show"
        number_sections: TRUE
        toc: true
        toc_depth: 4
        toc_float: true
        theme: flatly
        highlight: tango
        css: ../rmd_style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      message = FALSE
)
```

> Main packages used:`purrr`, `base R`\
> Main functions covered: `for`, `if`, `else`, `while`, `function(x){}`, `apply`, `lapply`, `purrr::map`  


> **Supplementary resources:** 
> 
> - [purrr cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/purrr.pdf)



# Loops

This session focuses on ways to **save time and keystrokes**: writing functions and iterations. As a rule of thumb if you find yourself writing (copying) the same code more than twice then it should be a function or loop. This section will be more focused on base R solutions, so we will write loops, then embed our loops in functions. Then, we will discuss two equivalent to writing loops: the `apply` function family from base R, and the `purrr:map` family from the `tidyverse`.

Loops get a bad rep in the R circles because they tend to be slower than the vectorized equivalents. However, the speed difference is not that earth shattering given the recent developments of R so if you want to write loops then don't let people throw this "But R loops are slow!" fallacy at you.
  
  
Loops are a way of iterating a given operation over a set of different inputs. We start by loading our packages and a subsetted msleep data. 

```{r, echo=FALSE}
# !diagnostics off
```


```{r message=FALSE, warning=FALSE}
library(dplyr)
library(purrr)
library(ggplot2)
```

Our data we will be using.
```{r}
msleep_df <- msleep %>% 
    select(name, sleep_total, sleep_rem, awake)

set.seed(2020) # this will ensure that our random numbers will be the same
```

## For loops

A for loop looks like:

```{r eval=FALSE}
for (value in that) {
this
}
```

What a for loop essentially does is that for every `value` in the `that` vector/object, do the `this` operation. It's best to explore this via a bare bones exercise. The following  script gives you a code, where we take every element (`i`) (but it could be named anything really) in the `1:5` vector and have the `print()` function print out the same message five times.


```{r}
# simple for loop
for (i in 1:5) {
    print("Hey, you are doing great!")
}

```

We can set up a loop to perform a set of operations on our input vector and put the result in a pre-specified output vector. Let's take the msleep data and compute the mean value for each column. First, lets select the numerical variables from the mammals sleep dataset.

```{r}
msleep_num <- msleep_df %>%
    select_if(is.numeric)
```


We want to compute the mean for every column. Instead of writing out `ms_colmean[1] <- mean(ms_colmean$var, na.rm = TRUE)` 3 times, changing the variable and row index we can write a nice loop in three lines. (imagine if we would have 15 numerical variables, we would still need this three lines of code.) The `sec_along` function defines that we want to loop over the three columns in our data.

```{r error=TRUE, warning=TRUE}
ms_colmean <- vector("double", length(msleep_num))
                     
for (i in seq_along(msleep_num)) {
    ms_colmean[i] <- mean(msleep_num[i])
}
```



Why did we get an error? (think about the difference between `[` and `[[`)

hint:
```{r}
msleep_num[2] 

# or

msleep_num[[2]]
```

Let's correct our indexing in the loop.
```{r}
ms_colmean <- vector("double", length(msleep_num))

                     
for (i in seq_along(msleep_num)) {
    ms_colmean[[i]] <- mean(msleep_num[[i]])
    print(ms_colmean[[i]])
}

```


**If we experience any problems with our loops, let's try deconstructing them piece by piece to understand their building blocks.**

First, what are we iterating on? A numerical vector
```{r}
seq_along(msleep_num)
```

Second, what happens inside the loop? We calculate a column mean (first, second, third)
```{r}
mean(msleep_num[[1]])
```


Third, we are assigning this value to the `ms_colmean` object (a numerical vector in this case)

```{r}
ms_colmean[[1]] <- mean(msleep_num[[1]])
```

Lastly, we print out the assigned value.
```{r}
print(ms_colmean[[1]])
```

And the loop does this whole thing three times in a row.


It seems that we still need to get rid of the `NA` problem. The solution is to use the `na.rm = TRUE` option when we compute the mean.

```{r}
for (i in seq_along(msleep_num)) {
    ms_colmean[[i]] <- mean(msleep_num[[i]], na.rm = TRUE)
    print(ms_colmean[i])
}
```



Loops can be nested in each other as well. To demonstrate this, we will do a multiplication table (a `10x10` matrix) with a nested for loop.

First we create an empty matrix that we will populate.

```{r}
mult_table <- matrix(NA, nrow = 10, ncol = 10)

mult_table
```


Then we write the loop that populates the empty matrix. We nest a second loop inside the outer one so our numerical vector will create a proper multiplication table at the end.

```{r}
num1 <- 1:10 # our input vector

for (i in num1) {               
    for (j in num1) {           
        mult_table[i,j] <- i*j 
    }
}

mult_table
```



## Flow controls

You have to control the Flow, Luke!

### If, else

It is a conditional statement, which we can put in a for loop if we want.

```{r eval=FALSE}
if (this) {
Plan A
} else {
Plan B
}

```

A quick example(s). 
First, we adapt our first loop for the `if` conditional statement. It prints out different values depending on if our statement is `TRUE` or `FALSE`.

```{r}
x <- 5

if (x > 5) {
    print("This is greater than 5")
} else {
    print("this not as great as 5 :(")
}
```



A more complicated example using multiple if conditions, where **we are curious if the input is even or odd**. We also want a nicer output with more communication. Again, let's build the loop step by step!

To achieve this, we modify our loop somewhat. We build in a stop condition, to check if our input is the one that we need. Since we only care about integers, we will put this conditional statement into our loop. The `stop` function will stop the loop from iterating whenever the loop arrives at that point where the `if`  statement is `TRUE` (in this case).

```{r eval=FALSE}
stop_cond <- is.integer(input)
     if (!stop_cond) {
         stop("input must be integer!")
     } else { rest of our loop}
```


Next we come up with some way to test if a number is even or odd. We will use the `%%` (modulus) operator that returns the remainder from the division, which in case of x/2 should be zero if x is even.

```{r}
42 %% 2

133 %% 2
```

The matching conditional would be something like this. In this code, the `cat` functions prints out our number in question and attaches the character string after the comma.

```{r eval=FALSE}
num_test <- input[i]%%2 == 0 
         if (num_test) {
             cat(i, "is even; ")
         } else {
             cat(i, "is odd; ")
         }
```


Finally, we have all the necessary building blocks, so let's put our loop together!

```{r}
input <- c(1:7)

for (i in seq_along(input)){
     stop_cond <- is.integer(input)
     if (!stop_cond) {
         stop("input must be integer!") # given the nature of our test condition, this only works on integers
     } else {
         num_test <- input[i]%%2 == 0 # the `%%` operator returns the modulus. if 0 then even, if not, odd.
         if (num_test) {
             cat(i, "is even; ") # cat() is a more flexible print function, which can combine objects and strings
         } else {
             cat(i, "is odd; ")
         }
     }
 }
```

Our error message is also working as intended!

```{r error=TRUE}
input <- c(1.5, 2, 4)

for (i in seq_along(input)){
     stop_cond <- is.integer(input)
     if (!stop_cond) {
         stop("input must be integer!") # given the nature of our test condition, this only works on integers
     } else {
         num_test <- input[i]%%2 == 0 # the `%%` operator returns the modulus. if 0 then even, if not, odd.
         if (num_test) {
             cat(i, "is even; ") # cat is a more flexible print function, which can combine objects and strings
         } else {
             cat(i, "is odd; ")
         }
     }
 }
```


A more practical example on using conditional statement is to recode a variable. We will create a new column in our curtailed `msleep_df` data frame and fill it with `NA`'s initially. Then we will create a dummy variable, which is 1 if `sleep_value > 1` AND `awake_value > 18` and 0 if this condition is not met.

```{r}
# if else condition inside the loop
msleep_df$new_awake <- NA

for (i in 1:nrow(msleep)) {
    sleep_value <- msleep_df$sleep_total[i]
    awake_value <- msleep_df$awake[i]
    
    test <- sleep_value > 1 & awake_value > 18
    
    if (test) {
        msleep_df$new_awake[i] <- 1
    } else {
        msleep_df$new_awake[i] <- 0
    }
}

msleep_df$new_awake
```


> Quick excercise:
> Write a loop which takes in a character vector and prints out "IT'S A CAT" if the value of the string is "cat" and "It's not a cat :'(((" whenever the value is not "cat". <br>
> The result should look something like this for the following input `x <- c("cat", "mouse")`:

```{r echo=FALSE}
x <- c("cat", "mouse")


for (i in 1:length(x)) {
  is_string <- is.character(x)
  if (!is_string) {
    stop("The input needs to be a character vector.")
  } else {
    if ((x[[i]] == "cat")) {
      print("IT'S A CAT")
    } else {
      print("Its not a cat :'(((")
    }
  }
}
```


### While

In each iteration of the loop, the `while` condition must be updated according to a certain logic, and it is evaluated again at the beginning of the next cycle of the loop to decide whether to go through or to stop. 

```{r eval=FALSE}
while (condition){
  # Do whatever is here as long as the condition is TRUE. 
}
```


For illustrative purposes, let's rewrite our first little for loop!


```{r eval=FALSE}
i <- 0 # set our initial value

while (i < 5 ) {
    print(i)
}
```

Press the `Esc` to stop our infinite loop! What just happened? We need to ensure that at one point, our while condition is met so our loop ends.

```{r}

while (i < 5 ) {
    print(i)
    i <- i+1 # this adds +1 to our `i` which then will reach 5 and stop our loop.
}
```


# Functions

For functions the same logic applies: if you have to copy paste/write the same line twice, think of a way to turn it into a function. The syntax of the `funtion()` function is the following:

```{r eval=FALSE}
name <- function(variables) {
    this is where we define our function. 
}
```

As with loops, you need to be consistent within your function with the naming of various interim objects, inputs and outputs. To get a feel for creating a function, let's create one, which will exponentiate a chosen base to our chosen exponent.

```{r}
my_power <- function(base, exp){
    output <- base ^ exp
    return(output)
}

# check out our function
my_power(base = 2, exp = 6)
```

Conditional statements within the functions work according to the same logic as in the loops discussed previously. We should add a much needed error message to our function:

```{r}
# add some error messages to our function with an if else + stop combination
my_power2 <- function(base, exp){
    cond <- is.numeric(base)
    if (!cond) {
        stop("base must be numeric!")
    } else {
        output <- base ^ exp
        return(output)
    }
}
```

```{r error=TRUE, collapse=FALSE}
my_power2("2", 4)

# we can experiment, as long as our inputs are numeric:
my_power2(2, 1:5)

my_power2(1:5, 2)
```

Or we can simulate a dice roll, with the use of the `sample()` function. If you are interested in how to build such simulations in R, you can check out *Grolemund, Garrett. Hands-On Programming with R: Write Your Own Functions and Simulations*

```{r}
# create a dice rolling function
roll <- function(){
    die <- 1:6
    dice <- sample(die, size = 1, replace = TRUE)
    
    return(dice)
}

roll()
```


It is now time to put our knowledge to good practical use and combine a function and loops. Since we calculate column means frequently, we should just probably write a function for it.

```{r}
column_mean <- function(df) {
    output <- vector("double", length(df))
    for (i in seq_along(df)) {
        output[[i]] <- round(mean(df[[i]], na.rm = TRUE),2)
    }
    
    return(output)
}

column_mean(msleep_df)
```

If we want to have a more general summary function, we can supply a function as argument to our function.

```{r collapse=FALSE}
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[[i]] <- fun(df[[i]], na.rm = TRUE)
  }
  return(out)
}

col_summary(msleep_df, sd)

col_summary(msleep_df, median)

```


# `apply` function family

There are several arguments against loops in the R community. The main one is that they used to be slow (that changed in recent years). There are many task that you can use vectorized operations in R which are usually a lot faster than the looped solutions. For example we want to add together two vectors, `vector1` and `vector2`. The loop solution is the following:

```{r collapse=FALSE}
vector1 <- c(0:10)
vector2 <- c(15:25)

vector3 <- NA

for (i in 1:length(vector1)) {
  vector3[i] <- vector1[i] + vector2[i]
}

vector3

```

The vectorized form:
```{r}
vector3_v <- vector1 + vector2

vector3_v
```



The `apply` family functions are usually take around the same time as a loop, but their advantage is that the syntax is easier and it is more compact code. In this section we'll go over `apply()`, `lapply()`, 

## `apply`

The `apply` function let's us apply a function to the rows or columns of our data frame or matrix by adjusting the `MARGIN = ` argument. **1 for row**, **2 for column**.

```{r}
df <- data.frame(x = rnorm(5),
                 y = rnorm(5),
                 z = rnorm(5))

df
```

```{r}
# sum over each column
apply(df, 2, sum)

# sum over each row
apply(df, 1, sum)

```

We can use it to check the number of missing values in our data frame as well (which is a very useful thing to do). here we "wrap" our function with the `function(x)`, otherwise we'll get an error. When in doubt, you can add `function(x)` even when it is redundant it won't cause any problem

```{r}
apply(msleep_num, 2, function(x) sum(is.na(x)))

# redundant `function(x)`
apply(df, 2, function(x) sum(x))


```




## `lapply`

The `lapply` function slightly differs from the `apply`:  
* It takes two arguments: `lapply(list, function)`
* It iterates the function over vectors or lists. This means that our output will also be a list.

Let's see what happens when we put a data frame into `lapply`. As a data frame is essentially lists (as columns) put together we get a result for each column (as a list). If we want a vector, we need to embed our `lapply()` function in an `unlist()`.
```{r}
# output as list
lapply(df, sum)

# output as vector
unlist(lapply(df, sum))
```

We can use it to create a list where every element is a matrix.

```{r}
mat_out <- lapply(1:3, function(x) matrix(x, nrow = 5, ncol = 5))

mat_out
```




# `purrr` package for iteration

Another way of iterating over our data is the `map` function from the `purrr` package. You can specify what sort of output you want after the `map_` part.

```{r}
map_dbl(df, mean)

map_dbl(mat_out, mean)
```

The `map` function can be part of a pipe chain. A small example of that looks like this

```{r}
msleep_df %>% 
    select_if(is.numeric) %>% 
    map_dbl(mean, na.rm = TRUE)
```



