---
title: "Coding challanges for Session Seven"
subtitle: ""
author: "Akos Mate"
date: '2021 March'
output:
    html_document:
        code_folding: "show"
        number_sections: FALSE
        toc: true
        toc_depth: 4
        toc_float: true
        theme: flatly
        highlight: tango
        df_print: paged
        css: ../rmd_style.css
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      message = FALSE,
                      warning = FALSE
)
```


```{r echo=FALSE}

library(tidyverse)
library(gapminder)

```

## t-test
> Load the palmerspenguins dataset and see if the difference between the bill lenght between the Adelie and Chinstrap are statistically significant or not. Tip: use the previously learned filter() function from dplyr to prep the data. You should get something similar as below. You can turn the scientific notation off with `options(scipen = 999)`.

```{r echo=FALSE}
penguins_t_prep <- palmerpenguins::penguins %>% 
    filter(species == c("Adelie", "Chinstrap"))

penguins_t <- t.test(bill_length_mm ~ species, data = penguins_t_prep)

options(scipen = 999)

penguins_t
```



# regression
> Using the `gapminder` dataset, filter our observations to the year of 2007 and see what are the effects of the GDP per capita, population and continent on the life expectancy. 

The result should be something like this. How would you interpret this?

```{r echo=FALSE}
data(gapminder)

options(scipen = 5)

gap_prep <- gapminder %>% 
    filter(year == 2007)

reg_gap <- lm(lifeExp ~ gdpPercap + pop + continent,
          data = gap_prep)
```

```{r}

summary(reg_gap)
```
